# Introducing Bayesian statistics

Before we turn to the use of Bayesian latent class models for estimating diagnostic test accuracy and disease prevalence, it is important to review several key principles of Bayesian statistics. The aim of this chapter is not to provide a comprehensive treatment of Bayesian methods; many excellent textbooks already serve this purpose. Rather, our objective is to present the core concepts that will be most useful for understanding and applying Bayesian approaches in the context of latent class modeling.

In this chapter, we first present Bayesian statistics and discuss the main differences and similarities between Bayesian statistics and the more commonly taught frequentist approach. We introduce the fundamental components of a Bayesian analysis, including prior distributions, likelihoods, and posterior inference. Finally, we describe how Bayesian models are typically fitted using Markov chain Monte Carlo (MCMC) methods.

## Bayesian Statistics

The term *Bayesian statistics* originates from the work of Reverend Thomas Bayes (1701–1761), and more specifically from Bayes’ theorem. Bayes’ theorem is a fundamental result in probability theory that describes the relationship between the probabilities of two events.

If *A* and B are two events, Bayes’ theorem states that:

$$
P(A \mid B) = \frac{P(B \mid A) \cdot P(A)}{P(B)} 
$$ {#eq-bayes}

Bayes’ theorem states that the probability of event *A* given that event *B* has occurred depends on three quantities: the probability of observing *A*, the probability of observing *B* if *A* has occurred, and the overall probability of observing *B*. It is therefore a simple but powerful law of probability. It can be applied to a wide range of epidemiological problems, and predictive values are, in fact, a direct application of Bayes’ theorem.

To illustrate this, let *A* denote the event of being diseased (D+) and B the event of testing positive (T+). The probability of disease conditional on a positive test result, P(D+\| T+), is referred to as the **positive predictive value** (PPV):

$$
P(A \mid B) = P(D^+ \mid T^+) = PPV 
$$ {#eq-ppvbayes1}

The probability of testing positive given that an individual is diseased is known as the **sensitivity** of the test (Se):

$$P(B \mid A) = P(T^+ \mid D^+) = Se$$ {#eq-sebayes}

We denote the probability of being diseased, P(D+), as the **disease prevalence** (P):

$$P(A) = P(D^+) = P$$ {#eq-pbayes}

The probability of a positive test result is given by the sum of the probabilities of observing a **true positive** result and a **false positive** result:

$$
P(B) = P(T^+) = P(True^+) + P(False^+) = Se \cdot P + (1-Sp) \cdot (1-P)
$$ {#eq-ptposbayes}

Combining these elements using Bayes’ theorem yields the positive predictive value (PPV):

$$
PPV = P(D^+ \mid T^+) = \frac{P(T^+ \mid D^+) \cdot P (D^+)}{P(T^+)} = \frac{Se \cdot P}{Se \cdot P + (1-Sp) \cdot (1-P)}$$ {#eq-ppvbayes2}

### From Bayes’ Theorem to Bayesian Inference

Although Bayes’ theorem is a simple law of probability, Bayesian statistics goes well beyond this. In a broader sense, Bayesian statistics is a framework for statistical inference in which **prior knowledge and observed data are formally combined to produce updated knowledge**.

This parallel can be made explicit by reinterpreting the terms of Bayes’ theorem. In a Bayesian context, the quantity of interest is typically an unknown parameter, denoted $\theta$ (for example, a disease prevalence, a test sensitivity, or a regression coefficient). Bayes’ theorem can then be written as:

$$
P(\theta \mid data) = \frac{P(data \mid \theta) \cdot P(\theta)}{P(data)}
$$ {#eq-Bayesinferences}

Each term of this expression has a clear **inferential interpretation**:

-   $P(\theta)$ represents existing knowledge or beliefs about the parameter before observing the data. We will later see that this knowledge is not expressed as a single estimate, but rather as a *distribution* of plausible values, referred to as the **prior distribution**.

-   $P(data|\theta)$ is the **likelihood**, which quantifies how compatible the observed data are with different values of the parameter.

-   $P(\theta | data)$ is the **posterior distribution**, which represents updated knowledge about the parameter after combining the prior information with the data.

-   $P(data)$ is a normalizing term that is required mathematically but is not of direct interest for epidemiological interpretation and **can generally be ignored** in practice.

Thus, Bayesian inference can be viewed as a formal learning process: **prior information is updated by the data through Bayes’ theorem to yield posterior knowledge**.

This structure closely mirrors the example of predictive values. In that context, disease prevalence plays the role of prior information. Indeed, prior to applying a diagnostic test, the best population‑level estimate of the probability that an individual is diseased is typically the disease prevalence. Test sensitivity and specificity form the likelihood, as they describe how likely a positive or negative test result is, given the true disease status. Finally, the positive predictive value corresponds to a posterior probability. By combining initial knowledge about disease prevalence (**the prior**) with the observed test result (**the data**), knowledge is updated in the form of a predictive value (**the posterior**).

Bayesian statistics generalizes this logic beyond diagnostic testing, providing a unified framework for statistical inference across a wide range of epidemiological problems.

### Bayesian *vs.* Frequentist Inferences

Frequentist statistics traditionally form the core of most medical statistics curricula, whereas Bayesian statistics are often underrepresented or omitted. This imbalance stems largely from historical limitations: for many years, there were no practical computational “engines” capable of solving anything beyond very simple Bayesian problems, such as estimating a single proportion. As a result, Bayesian methods had few direct applications for statistical inference in practice.

This situation changed dramatically with the development of **Markov chain Monte Carlo (MCMC)** methods, particularly **Gibbs sampling**, introduced in 1984 by Stuart and Donald Geman[@Geman1984]. These computational advances made it possible to perform Bayesian inference for complex models that had previously been intractable, thereby opening the door to widespread practical use of Bayesian statistics.

Before exploring these methods in detail later in the chapter, we first examine some of the key differences and similarities between the frequentist and Bayesian approaches to statistical inference.

In **frequentist inference**, an unknown parameter (for example, a disease prevalence) is treated as a **fixed but unknown quantity**. The parameter is assumed to have one fixed true value, which does not vary, even though it is not known. In contrast, the observed data are regarded as **random**, representing one of many possible datasets that could have been observed under repeated sampling. Inference proceeds through a **deductive approach**, asking the question: *what is the probability of observing these data, given a specified hypothesis or parameter value?*

In **Bayesian inference**, the unknown parameter is instead regarded as a **random quantity** and is represented by a probability distribution, with some values more plausible than others. The data are treated as **fixed** once observed and are used to update prior knowledge about the parameter. Inference follows an **inductive approach**, addressing the question: *what is the probability of the hypothesis or parameter value, given the observed data?*

The distinction between deductive and inductive approaches to inference is illustrated below (@fig-deduindu), using a medical example. In this example, we may imagine a physician awaiting the next patient. Before seeing the patient, the physician could formulate a hypothesis regarding the disease affecting the patient (A, B, or C), corresponding to a null hypothesis in the frequentist framework. The physician would then collect data on the patient’s symptoms and evaluate the probability of observing these symptoms given the initial disease hypothesis. If this probability were sufficiently low—below a prespecified threshold α (the type I error rate)—the physician might reject the initial hypothesis. This reasoning exemplifies a deductive approach to inference, proceeding from a hypothesized cause to expected observations.

A more commonly encountered approach in medical inference is inductive reasoning. In this case, the physician does not begin by formulating an explicit hypothesis. Instead, the physician first examines the patient and collects information on the observed symptoms. Based on these observations, the physician then assesses the relative plausibility—or probability—of diseases A, B, or C. In this inductive framework, inference proceeds from the data to conclusions about the underlying cause, reflecting a Bayesian approach.

```{r fig-deduindu}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Parallels between deductive and inductive processes in medical inference. 


library(ggplot2)

# -----------------------------
# Create conceptual data
# -----------------------------
df <- data.frame(
  x = c(rep(0, 3), rep(1, 5)),
  y = c(3:1, 4:0),
  label = c(
    "Disease A", "Disease B", "Disease C",
    "Fever", "Cough", "Rash", "Swelling", "Pain"
  ),
  type = c(rep("Disease", 3), rep("Symptom", 5))
)

# -----------------------------
# Plot
# -----------------------------
p <- ggplot(df, aes(x, y)) +
  geom_text(
    aes(label = label),
    size = 4
  ) +
  geom_segment(
    aes(x = 0.1, xend = 0.9, y = 4.5, yend = 4.5),
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  geom_segment(
    aes(x = 0.9, xend = 0.1, y = -1, yend = -1),
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  annotate("text", x = 0.5, y = 4.9, label = "Deduction (Frequentist)", size = 4.5) +
  annotate("text", x = 0.5, y = -1.4, label = "Induction (Bayesian)", size = 4.5) +
  annotate("text", x = 0, y = 5.6,
           label = "Putative underlying diseases",
           fontface = "bold", size = 4.5) +
  annotate("text", x = 1, y = 5.6,
           label = "Observed symptoms",
           fontface = "bold", size = 4.5) +
  scale_x_continuous(limits = c(-0.4, 1.4)) +
  scale_y_continuous(limits = c(-2, 6)) +
  theme_void() +
  theme(
    plot.margin = margin(20, 20, 20, 20)
  )

print(p)
```

Despite these fundamentally different approaches to inferences, there are nevertheless many similarities between frequentists and Bayesian inferences.

For instance, **both approaches can be used to address similar types of problems**, such as:

-   estimating a mean (for a quantitative variable);

-   estimating a proportion (for a binary variable);

-   comparing two means;

-   comparing two proportions;

-   estimating coefficients in linear, logistic, or more generally, generalized linear models;

-   and many others.

In both frameworks, the **unknown parameter is typically summarized by a point estimate** (e.g., a population prevalence of 15%), and **uncertainty is quantified using an interval estimate** (e.g., from 10% to 20%). The interpretation of a frequentist 95% confidence interval, however, is often subtle and difficult to communicate correctly. The proper interpretation is as follows: *if the study were repeated an infinite number of times, and a 95% confidence interval were computed for each repetition, then 95% of these intervals would contain the true parameter value*.

Importantly, this interpretation does not apply to any single observed interval. For a given dataset, we do not know whether the computed interval contains the true parameter value or not. For example, in @fig-95ci, if Sample 6 were the observed dataset, the corresponding frequentist 95% confidence interval would fail to include the true prevalence of 15%.

```{r}
#| message: false
#| warning: false
#| include: false

set.seed(123)

n_samples <- 20
n_per_sample <- 200
true_prev <- 0.15

dat <- do.call(rbind, lapply(1:n_samples, function(i) {
  y <- rbinom(n_per_sample, 1, true_prev)
  p_hat <- mean(y)
  se <- sqrt(p_hat * (1 - p_hat) / n_per_sample)
  
  data.frame(
    sample = i,
    estimate = p_hat,
    lower = p_hat - 1.96 * se,
    upper = p_hat + 1.96 * se
  )
}))
```

```{r fig-95ci}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Estimated proportions and frequentist 95% confidence intervals from 20 samples of 200 participants drawn from the same population with a true disease prevalence of 15%.

library(ggplot2)

ggplot(dat, aes(x = sample, y = estimate * 100)) +
  # Confidence intervals
  geom_errorbar(
    aes(ymin = lower * 100, ymax = upper * 100),
    width = 0.15,
    color = "black"
  ) +
  # Point estimates
  geom_point(
    aes(color = factor(sample)),
    size = 2.5
  ) +
  # True prevalence
  geom_hline(
    yintercept = true_prev * 100,
    linetype = "dashed",
    linewidth = 0.6
  ) +
  scale_x_continuous(breaks = 1:n_samples) +
  labs(
    x = "Sample",
    y = "Prevalence (%)"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12, face = "bold")
  )

```

The **Bayesian 95% credible interval (95% BCI)**, by contrast, is constructed from the **posterior distribution of the unknown parameter**. Recall that this posterior distribution is obtained by combining prior knowledge about the parameter (for example, evidence from an earlier study suggesting a prevalence of 12% in the population) with the observed data (for example, 30 diseased individuals in a sample of 200 participants).

The interpretation of a 95% BCI is straightforward: *given the observed data and the specified prior, there is a 95% probability that the true parameter value lies within this interval*. This probability statement directly reflects uncertainty about the parameter itself.

It is important to note that this interpretation—although correct for Bayesian credible intervals—is **often incorrectly applied to frequentist 95% confidence intervals** in the scientific literature.

Some important advantages of Bayesian inference, compared with frequentist inference, lie in its ability to model **latent variables**. A latent variable is a variable that cannot be observed directly or was not measured in a given study. When such a latent variable can be linked to observed data through a statistical model, a Bayesian approach can offer a solution. As expected, this type of problem will be central to several subsequent chapters of this book.

Bayesian inference also offers important advantages in its ability to **produce 95% Bayesian credible intervals for a wide range of parameters**. As discussed in @sec-reporting-results, many derived quantities—such as likelihood ratios, predictive values, or misclassification cost terms—do not have a single, universally accepted closed‑form frequentist confidence‑interval formula. In the Bayesian framework, however, these quantities are treated no differently from any other parameter: a posterior distribution can be obtained for each of them, and a corresponding 95% BCI can therefore be computed directly.

Bayesian inference also offers additional advantages over frequentist inference, although these are not central to the present discussion. For example, Bayesian methods are generally more robust to data that are missing at random, in contrast to the complete‑case analysis that is often used by default in frequentist approaches. Bayesian inference also provides a flexible and often straightforward framework for modeling complex data structures, as well as for correcting estimates from epidemiological studies affected by misclassification bias, measurement error, or residual confounding.

On the other hand, **a potential disadvantage** of Bayesian inference is that the **resulting inferences depend on the specification of prior distributions**. If these priors are poorly specified or incorrectly elicited, the resulting posterior distribution for an unknown parameter may be misleading or, in some cases, improper. Such issues are more likely to arise in studies with small sample sizes, particularly when overly narrow or informative prior distributions are used. In these situations, the limited amount of new data contributes relatively little when combined with prior information to form the posterior distribution.

Nevertheless, because Bayesian inference inherently depends on prior assumptions, it is essential to assess the robustness of results to these assumptions. As we will discuss later, **conducting sensitivity analyses using alternative or perturbed prior distributions is therefore a critical step** to evaluate how sensitive the conclusions are to the choice of prior.

### Main Elements of a Bayesian Analysis

Regardless of the problem under consideration, a Bayesian analysis is always built around three core components:

-   **a** **prior distribution** for each unknown parameter;

-   **a likelihood function**, which describes how the observed data are generated given the unknown parameters;

-   **a posterior distribution** for each unknown parameter, representing updated beliefs about the parameters after observing the data.

In practice, a complete Bayesian analysis will also typically include one or more sensitivity analyses to assess the impact of different prior specifications on the resulting inferences. In the remainder of this section, however, we focus on the first three core elements of a Bayesian analysis, which form the foundation of Bayesian inference. Sensitivity analyses, although essential for evaluating the robustness of results to prior assumptions, are discussed separately in a later chapter.

#### Prior Distributions

A distinguishing feature of Bayesian analysis is the explicit use of a **prior distribution** for each unknown parameter. This probability distribution summarizes information about the parameter that originates from sources **external to the observed data**. The notion of incorporating external information is central to the Bayesian framework.

Importantly, the observed data themselves **must not** be used to construct the prior distribution that will later be combined with those same data. Doing so would amount to using the data twice, which undermines the interpretation of the posterior distribution. Instead, prior distributions are intended to represent uncertainty about the parameter **before** observing the data, and this uncertainty is expressed formally through a probability distribution.

Whenever possible, prior distributions should be based on the most **reliable and relevant external sources of information** available. For example, when estimating the prevalence of a disease in a population, plausible sources for constructing a prior distribution may include:

-   a **peer‑reviewed meta‑analysis** synthesizing data from multiple studies conducted in a similar population (e.g., same geographical region and time period) and using comparable case definitions (although such sources are often unavailable);

-   one or several **peer‑reviewed studies** with characteristics similar to those of the study being conducted;

-   **non‑peer‑reviewed data**, such as governmental or surveillance reports;

-   **expert opinion**;

-   or, in the absence of stronger evidence, a **subjective belief**, for example: *it would be highly unlikely for the disease prevalence to exceed 50%*.

It is also possible to specify a ***vague*** **prior distribution** when limited external information is available. For instance, a **uniform prior** assigning equal probability across the entire plausible range of values can be used. In the case of disease prevalence, this could correspond to a uniform distribution between 0% and 100%, explicitly stating that all values within this range are considered equally plausible.

Historically, such priors were sometimes referred to as *non‑informative priors*. This terminology can be misleading, however, since assigning equal probability to all values still represents a form of information. A uniform prior does not indicate the absence of prior beliefs; rather, it reflects a deliberate choice to express maximal uncertainty within a defined range.

For example, if a surveillance report conducted in the same region a few years earlier estimated a disease prevalence of 30%, with a 95% confidence interval ranging from 13% to 48%, the distribution illustrated below (@fig-betaPrev) could be used to represent our prior beliefs about the disease prevalence.

```{r fig-betaPrev}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Density curve of a prior distribution based on a mode of 30% and a upper 97.5th percentile of 48%.


library(epiR) 

prev <- epi.betabuster(mode=0.30,                   
                       conf=0.975, 
                       imsure="less than", 
                       x=0.48,
                       )  

curve(dbeta(x, 
            shape1=prev$shape1, 
            shape2=prev$shape2), 
      from=0, to=1,
      xlab = "Prevalence", 
      ylab = "Density")
```

If no external source of data is available, this uncertainty can instead be represented explicitly by using a distribution that assigns equal probability to values between 0% and 100% (i.e., a vague prior distribution; @fig-vaguePrev).

```{r fig-vaguePrev}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Density curve of a prior distribution assigning equal probability to values between 0 and 1 (i.e., 0 to 100%).

curve(dbeta(x, 
            shape1=1.0, 
            shape2=1.0), 
      from=0, to=1, 
      xlab = "Prevalence", 
      ylab = "Density") 
```

In a later section, we review the different types of distributions that can be used to model various kinds of parameters. The choice of a prior distribution should always be consistent with the nature of the parameter of interest. A detailed discussion of commonly used distributions is therefore deferred to that section.

#### The Likelihood Function

The likelihood function describes how the observed data are linked to the unknown parameters of the model. In this respect, it closely resembles the frequentist framework. For example, in a frequentist analysis, a binary outcome may be related to an unknown intercept and the unknown effect of a predictor through a logit link, as in a logistic regression model. In Bayesian analyses, however, likelihood functions are typically stated explicitly and formally, whereas in many frequentist studies they are sometimes only mentioned implicitly or described at a higher level.

The likelihood function acts as the engine through which information contained in the data contributes to inference about the unknown parameters, in combination with the prior distributions. As a result, the form of the likelihood function depends strongly on the problem under consideration and on how the data were generated.

For example, consider a situation in which the outcome of interest can take only two possible values—healthy or diseased (coded as 0 or 1). In this case, we may wish to estimate an unknown parameter such as the proportion of diseased individuals in the population (*P*, the disease prevalence) based on the observed data: the number of positive individuals (*T*) and the total number of individuals tested (*n*).

For instance, if 30 individuals are observed to be diseased (*T*=30) among 100 tested individuals (*n*=100), the unknown parameter *P* can be linked to the data through a simple likelihood function, as described below.

$$T \sim dBin(P, n)$$ {#eq-binomialPrev}

In words, this means that the number of positive individuals (*T*) is assumed to follow a binomial distribution (denoted here as *dBin*), characterized by two quantities: *P*, which represents the disease prevalence, and *n*, the number of individuals tested. This notation can initially be disconcerting for readers new to Bayesian modeling. Indeed, in this expression, the observed data (*T* and *n*) appear on both sides of the equation, while the unknown parameter of interest (*P*) is embedded within a distribution that also includes one observed quantity (*n*). Nevertheless, once values for *T* and *n* are specified and a prior distribution for *P* is chosen, this model can be used to derive a posterior distribution for *P*.

This formulation reflects a general feature of many Bayesian likelihood functions. There is not necessarily a natural left‑to‑right (or right‑to‑left) reading of the equation. Instead, the likelihood must be interpreted as a whole, as a mathematical statement describing how the observed data are generated given the parameters. In more complex models, this can make it somewhat difficult to disentangle observed data from unknown parameters, particularly for readers encountering Bayesian notation for the first time.

As another classic example, @eq-logistic below presents the likelihood function for a logistic regression model estimating two unknown parameters: an intercept ($\beta_0$) and the effect of an exposure ($\beta_1$) on the log‑odds of disease, based on the observed outcome (*y*) and exposure (*exp*) values recorded for multiple individuals.

$$
\begin{aligned}
& y \sim \text{dBern}(P) \\
& \text{logit}(P)=\beta_0 + \beta_1 \cdot exp
\end{aligned}
$$ {#eq-logistic}

In words, the outcome *y* is assumed to follow a Bernoulli distribution, meaning that it can take only the values 0 or 1, with probability *P* of taking the value 1. The logit transformation of this probability is modeled as a linear function consisting of an intercept ($\beta_0$) and a regression coefficient ($\beta_1$), the latter being multiplied by the value of the predictor *exp*.

In this formulation, the variables *y* and *exp* represent the observed data, whereas the intercept ($\beta_0$) and the coefficient ($\beta_1$) are the unknown parameters to be estimated. Note that, in this example, *P* is neither estimated directly as a model parameter nor observed as data. Instead, it acts as an intermediate quantity within the likelihood function, linking the observed data (*y* and *exp*) to the unknown parameters ($\beta_0$ and $\beta_1$).

#### Posterior Distributions

Posterior distributions represent our uncertainty about unknown parameters *after* observing the data. They describe how our prior beliefs are updated in light of the observed evidence. Like priors, posterior distributions are expressed as probability distributions and typically reflect a compromise between the prior information and the contribution of the data.

**In principle, Bayes’ theorem is used** to update prior distributions with observed data **to obtain posterior distributions** for unknown parameters. If *x* denotes the observed data, $f(\pi)$ the prior distribution of an unknown parameter $\pi$, and $f(x|\pi)$ the likelihood function, the posterior distribution, $f(\pi|x)$ is given by:

$$f(\pi|x) = \frac{f(x|\pi) \cdot f(\pi)}{\int_\pi f(x|\pi) \cdot f(\pi) \cdot d\pi}$$

**In practice, however, evaluating this expression analytically is usually not possible.** In the following sections, we describe how Markov chain Monte Carlo (MCMC) methods, such as Gibbs sampling, can be used to draw values directly from the posterior distribution without computing it in closed form. When a sufficiently large number of samples has been generated, posterior summaries can be obtained: for example, the median may be reported as a point estimate, while the 2.5th and 97.5th percentiles can be used to construct a 95% Bayesian credible interval. Alternatively, the posterior distribution itself may be visualized using a density plot.

Let us work through a complete example to illustrate how posterior distributions are generated.

**The problem.**\
We aim to estimate the prevalence of a given disease.

**The prior.**\
A surveillance report conducted in the same region a few years earlier estimated the disease prevalence at 30%, with a 95% confidence interval ranging from 13% to 48%. This prior distribution is illustrated in @fig-betaPrev.

**The observed data.**\
Thirty individuals were tested (we acknowledge the limited sample size), and 15 of them were found to be diseased.

**The likelihood function.**\
To estimate a prevalence parameter (*P*) based on the number of tested individuals (*n*) and the number of diseased individuals (*T*), we can use the likelihood function defined in @eq-binomialPrev .

By combining these elements, we can generate a **posterior distribution** for the unknown parameter *P*, representing the disease prevalence. Although we defer the technical details of this process to subsequent chapters, the purpose of this example is simply to illustrate how prior information—here represented by a distribution with a mode of 30% and a 95% confidence interval from 13% to 48%—can be **updated with new data** (15 diseased individuals out of 30 tested) to produce a posterior distribution.

Figure @fig-PriorPost illustrates how the prior distribution for disease prevalence is updated by the observed data to yield the posterior distribution, which represents our updated belief about the disease prevalence after incorporating the evidence provided by the data.

```{r fig-PriorPost}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-cap: Density curves of the disease‑prevalence prior (dashed red line) and posterior (solid black line), together with the point estimate based on the data alone (dashed blue line).

#Creating a dataset
datalist <- list(T=15, n=30)

# Specify the model
model_single_prop <- paste0("model{

#=== LIKELIHOOD ===#

  T ~ dbin(P, n)


#=== PRIOR  ===#

  P ~ dbeta(",prev$shape1,", ",prev$shape2,")    ## Prior for P

}")

#write as a text (.txt) file
write.table(model_single_prop, 
            file="Models/chp 3/model_single_prop.txt", 
            quote=FALSE, 
            sep="", 
            row.names=FALSE,
            col.names=FALSE)

#Initializing values for the parameter "P" for the 3 chains
inits <- list(
  list(P=0.10),
  list(P=0.50),
  list(P=0.90)
  )

library(R2jags)
library(coda)
#Run the Bayesian model
bug.out <- jags(data=datalist,                             
               model.file="Models/chp 3/model_single_prop.txt",         
               parameters.to.save=c("P"),               
               n.chains=3,                                 
               inits=inits,                                
               n.iter=10000,                                
               n.burnin=1000,                              
               n.thin=1,                                   
               DIC=FALSE)                                   

#Compute median and 95%BCI
samples_P <- bug.out$BUGSoutput$sims.list$P
qs <- quantile(samples_P, probs = c(0.025, 0.5, 0.975))
low  <- round(as.numeric(qs["2.5%"]), 3)
med   <- round(as.numeric(qs["50%"]), 3)
hi <- round(as.numeric(qs["97.5%"]), 3)


#Plot the posterior distribution
plot(density(x=bug.out$BUGSoutput$sims.list$P),          
     xlab="Prevalence", ylab="Density", 
     , main="")                 

curve(dbeta(x, 
            shape1=prev$shape1, 
            shape2=prev$shape2), 
      from=0, to=1,          
      lty=2,                    
      col="red",                
      add=TRUE)                


# Add vertical blue dashed line at x = 0.50
abline(v = 0.50, col = "blue", lty = 2)

```

This posterior distribution can be summarized by its **median**, equal to `r scales::percent(med)` and by its 95% BCI ranging from `r scales::percent(low)` to `r scales::percent(hi)`.

As anticipated, the posterior distribution lies **between** the prior distribution—characterized by a mode of 30%—and the value of *P* that would be estimated using the observed data alone (50%, corresponding to 15 diseased individuals out of 30). This illustrates how Bayesian inference combines prior information and data to yield an updated estimate of the disease prevalence.

#### Interplay Between Priors, Data, and Posteriors
