# Preface

Welcome to the first edition of *Inference Without a Gold Standard*. Writing this book was motivated by a long‑standing interest in a central challenge in epidemiology and health sciences: how to draw reliable inferences when no perfect diagnostic test exists.

Bayesian statistical methods are now widely used in both human and veterinary epidemiology, and among them, **Bayesian latent class models** have emerged as a particularly powerful framework for estimating disease prevalence and diagnostic test accuracy. These models are especially valuable **in situations where a so‑called *gold standard* test is unavailable**, unreliable, or impractical. Historically, the most commonly used diagnostic test was often treated as if it were a perfect reference when evaluating new tests or estimating prevalence. However, as diagnostic technologies continue to improve, this assumption is increasingly difficult to justify. In many cases, newer tests may be as accurate as—or even more accurate than—the tests used for comparison, leading to biased assessments when discrepancies arise.

Latent class approaches address this problem by treating the **true disease status of individuals as unobserved, or *latent***. Rather than classifying individuals as definitively infected or uninfected, the framework models uncertainty explicitly, combining observed test results with information about test performance and prior knowledge about disease prevalence and accuracy of the tests under investigation. Bayesian implementations of these models are particularly appealing, as they integrate prior information with data to produce posterior distributions that reflect updated beliefs about prevalence and test accuracy—without requiring knowledge of the true disease status of sampled individuals.

This book introduces these ideas in a practical and accessible way, to help readers apply Bayesian latent class models confidently and appropriately in real‑world settings.

## Who is this book for

This book is intended for readers interested in evaluating diagnostic tests or estimating the prevalence of a health condition in the absence of a perfect reference test. While many of the examples presented in the book are drawn from animal health, the methods are directly applicable to human health. Prior experience in animal health is therefore not required.

A basic understanding of epidemiological concepts related to diagnostic testing—such as gold standards, diagnostic sensitivity and specificity, and disease prevalence—will be helpful. The book is primarily aimed at graduate students who have completed introductory courses in epidemiology and statistics, as well as practicing epidemiologists and health researchers. Readers are **not expected to have prior training in Bayesian statistics**, as the necessary concepts are introduced progressively.

## What you will learn

In the first chapter, we review the key epidemiological parameters used to describe and compare the accuracy of diagnostic tests. We also discuss the main study designs that can be used to validate new diagnostic tests. This is followed, in the second chapter, by an introduction to Bayesian methods, their underlying principles, and a brief overview of their implementation in R.

We then move to the core focus of the book: estimating **true disease prevalence** and **diagnostic test accuracy** in the absence of a gold standard test. Building progressively in complexity, we consider situations in which:

-   **Two imperfect diagnostic tests** are applied to a single population;

-   Diagnostic tests may be **conditionally dependent**;

-   Two imperfect tests are applied to **two or more populations**;

-   **Three or more** **diagnostic tests** are applied to one or multiple populations;

-   Test sensitivity and specificity vary as a function of one or more **covariates**.

Following this extensive treatment of diagnostic test accuracy, we introduce **multistage models** for settings in which multiple individuals within a group (e.g., animals from a herd or individuals from different regions) are tested using imperfect diagnostic tests. These models allow inference on **group-level prevalence** and are particularly useful for assessing **freedom from disease**.

Finally, we consider settings in which a **single quantitative diagnostic test** is applied to a population comprising both normal and abnormal individuals. We show how **Gaussian finite mixture models** can be used to identify underlying subpopulations and to establish **reference intervals** for biological measurements when no clear diagnostic cutoff exists.

Throughout the book, we first present the **theoretical concepts** and then provide clear, **step-by-step guidance on how to conduct similar analyses using R**. At the end of each chapter, a series of exercises is provided to help readers consolidate their understanding. These exercises can be completed using the datasets and partially completed R scripts available in my [Github repository](https://github.com/dufours/Dx-LCM-Bookdown). Solutions are also provided in the form of fully completed R scripts.

## How to use this book

This book can be used in different ways. You may simply sit down and read it to gain a general overview of the methods. Even if you are not particularly interested in the R code, you should still gain a solid understanding of the approaches that can be used to estimate disease prevalence and diagnostic test accuracy when a gold standard test is unavailable.

That said, this book was primarily **designed for readers who want to apply these methods**. This can be done using the accompanying exercises, although you are also encouraged to use your own data alongside—or instead of—the proposed examples.

The book can also serve as the main **teaching material for a course** focused on the use of Bayesian latent class models for diagnostic test validation and prevalence estimation, for example at the graduate level following introductory courses in epidemiology. In fact, much of the material was originally developed to support an in‑person workshop that has been offered on multiple occasions. As a result, the content is organized in a logical sequence intended to facilitate the acquisition of methods that can sometimes be conceptually demanding.

Put differently, the material has been tested on multiple cohorts of guinea pigs. The first few groups may not have survived the experience, but the book is now presented in a *tested‑and‑approved* format. Still—let’s be honest—there is always room for improvement.

## Conventions

Throughout this document, you will find examples of **R code**, accompanied by explanatory comments. R code is always displayed in **grey boxes** (see the example below). This code can be reused or adapted for your own analyses.

Lines that begin with a \# symbol are **comments**. Comments are not executed by R; instead, they are used to explain what the code does.

Unless explicitly stated otherwise (i.e., "results are not shown"), each grey box containing R code is followed by the output generated by the analysis, such as numerical values, figures, or other results.

For example, the following R code displays the main **descriptive statistics** for the *speed* variable in the *cars* dataset. Note that this dataset is included by default in R.

```{r}
#This is a comment. R will ignore this line
summary(cars$speed)
```

Throughout this document, we will use the following conventions:

-   *Italics* are used for **datasets** or **variables**.\
    For example, the *speed* variable from the *cars* dataset.

-   **Shaded boxes** are used to display **R libraries** (e.g., `epiR`) and **R functions** (e.g., `epi.betabuster()`).

In R, you can use functions from a package in two equivalent ways:

1.  By first loading the library using `library()`, and then calling its functions directly.

2.  By calling a specific function using the package name followed by `::`, without loading the entire library.

For example, the following two pieces of code are equivalent:

```{r, eval=FALSE}
library(epiR)
Se <- epi.betabuster(mode=0.85, 
                     conf=0.975, 
                     imsure="greater than", 
                     x=0.75) 
##OR##
Se <- epiR::epi.betabuster(mode=0.85, 
                           conf=0.975, 
                           imsure="greater than", 
                           x=0.75) 

```

Both approaches produce the same result; the second is often used to make it explicit which package a function comes from. This practice may improve reproducibility, but it comes at the expense of longer code. Throughout this document, we will therefore always **load the library first** and then call its functions, in order to keep the code concise and easier to read.

You may also have noticed the line breaks within the `epi.betabuster()` function in the example above. These line breaks are not required for the code to run; however, they can improve readability by visually separating the different arguments passed to a function. This formatting style is particularly useful when a function has multiple arguments, and we will use it frequently throughout this textbook (an old habit!).

One final point concerns the naming of function arguments. In R, it is not mandatory to explicitly name all arguments, provided they are supplied in the order expected by the function.

For instance, the `epi.betabuster()` function expects the following arguments, in this order:

1.  a mode (`mode =`)

2.  a level of certainty (`conf =`)

3.  an indication of whether the specified value corresponds to a lower or upper percentile of the distribution (`imsure =`)

4.  the value itself (`x =`)

As a result, argument names can be omitted when they are provided in the correct sequence. The following two pieces of code are therefore equivalent:

```{r, eval=FALSE}
Se <- epi.betabuster(mode=0.85, 
                     conf=0.975, 
                     imsure="greater than", 
                     x=0.75) 
##OR##  
Se <- epi.betabuster(0.85, 
                     0.975, 
                     "greater than", 
                     0.75) 
```

Although omitting argument names can result in shorter and more compact code, explicitly naming arguments often improves clarity, particularly when working with unfamiliar functions or more complex syntax. For this reason, throughout this document we will consistently use the **fully specified version of the code**, in which all function arguments are explicitly named.

Because you are being introduced to these functions for the first time, relying on abbreviated syntax from the outset could make the code more difficult to interpret and follow. Using the longer, more explicit form is intended to facilitate understanding of each function’s structure and purpose. Once you have become more familiar with the functions and their expected arguments, you may choose to adopt the shorter syntax in your own work.

## Before you begin

Most chapters of this textbook are designed to encourage **active learning**. You are invited to copy and paste the R code provided throughout the chapters and to modify it in order to explore and experiment. In addition, most chapters include exercises intended to help you apply and consolidate the concepts and skills you acquire.

To fully engage with the material, you will need to install several software components before you begin.

### **R and RStudio**

First, install the most recent version of **R** from the [R Project](https://www.r-project.org/)'s website. R is a free software environment for statistical computing and graphics. This book was developed using **R version 4.4.3**.

Next, install **RStudio**, an integrated development environment for R. RStudio provides a user‑friendly interface that includes a console, a syntax‑highlighting editor with support for direct code execution, and tools for plotting, debugging, history management, and workspace organization. A free **RStudio Desktop Open Source** version is available from the [RStudio website](https://www.rstudio.com/products/rstudio/download/). This book was developed using **RStudio version 2026.01.1**.

### **Required R Packages**

We have made an effort to keep the number of required R packages to a minimum. For this textbook, you will need to install the following three packages:

-   **R2jags** (version 0.8‑9 was used).\
    The `R2jags` package provides a convenient interface between R and **JAGS**, a software program specialized in Bayesian analyses (see below). It automatically prepares the data and model files in a format that can be read by JAGS and facilitates running Bayesian models from within R.

-   **mcmcplots** (version 0.4.3 was used).\
    The `mcmcplots` package provides the function `mcmcplot()`, which produces commonly used Markov chain Monte Carlo (MCMC) diagnostic plots. These plots are written to an HTML file that can be viewed in a web browser, allowing efficient inspection of a large number of parameters by scrolling through the output.

-   **epiR** (version 2.0.91 was used).\
    The `epiR` package contains tools for the analysis of epidemiological and surveillance data. In this textbook, we will mainly use functions from `epiR` that derive distributional parameters based on expert elicitation (mainly the `epi.betabuster()` function). These parameters will be used to construct and visualize prior distributions.

### JAGS

Finally, install the most recent version of **JAGS** from [Sourceforge](https://sourceforge.net/projects/mcmc-jags/files/). JAGS is a free software package for performing Bayesian inference using Gibbs sampling. The name JAGS stands for *Just Another Gibbs Sampler*. It is a standalone program that implements MCMC simulation for Bayesian models and is used in conjunction with R via packages such as `R2jags`.

**Why JAGS?**

Several software platforms are available for Bayesian model fitting using MCMC methods, including **OpenBUGS** and **JAGS**. The primary motivation for using JAGS is **platform compatibility**. JAGS is actively maintained and provides stable, native support for **Windows, macOS, and Linux** systems. By contrast, OpenBUGS was originally developed for Windows, and its use on macOS often requires additional configuration or legacy solutions, which can present a barrier for Mac users. To ensure a smooth and consistent experience across operating systems, we therefore use JAGS in this textbook.
